The LBRY Trending Algorithm

### 2021-11-18 tex/ms.tex
Spike power

Our final choice of spike mass function is complex. I will describe simpler
versions at first, which will help in motivating the final, complicated, choice.
First, let's look at a choice where the LBC amounts are `softened' by raising
them to some power less than one (here I have chosen a power of 1/3, in
previous versions of the algorithm I used 1/4):
\begin{align}
M(\ell_t, \ell_{t+1}) &= \ell_{t+1}^{1/3} - \ell_t^{1/3}.
\end{align}
The size of a trending spike here would be the change in the `softened'
LBC amount. A change from 0 to 1 LBC would give a spike mass of 1,
as would a change from 1 to 1000 LBC, or 1000 to 1,000,000. It takes more
LBC to move a mountain than to move a molehill.

With this choice, it wouldn't matter whether a change took place in one large
step or in smaller increments\footnote{This would occur with any
choice of the form $M(x, y) = \Phi(y) - \Phi(x)$ for a monotonic function
$\Phi()$.}. Specifically,
\begin{align}
M(x, y) + M(y, z)
    &= y^{1/3} - x^{1/3} + z^{1/3} - y^{1/3} \\
    &= z^{1/3} - x^{1/3} \\
    &= M(x, z).
\end{align}
This may not be entirely desirable. Genuinely popular claims on LBRY tend
to receive several smaller supports, rather than one large one, and intuitively
this ought to help. A trending algorithm should show what's trending,
as well as what's recently been boosted.

More to come

>>>>

\subsection{Change in Softened LBC Amount}
Our final choice of spike mass function is complex. I will describe simpler
versions at first, which will help in motivating the final, complicated, choice.
First, let's look at a choice where the LBC amounts are `softened' by raising
them to some power less than one (here I have chosen a power of 1/3, in
previous versions of the algorithm I used 1/4):
\begin{align}
M_1(\ell_t, \ell_{t+1}) &= \ell_{t+1}^{1/3} - \ell_t^{1/3}.
\end{align}
The size of a trending spike here would be the change in the `softened'
LBC amount. A change from 0 to 1 LBC would give a spike mass of 1,
as would a change from 1 to 1000 LBC, or 1000 to 1,000,000. It takes more
LBC to move a mountain than to move a molehill.

With this choice, it wouldn't matter whether a change took place in one large
step or in smaller increments\footnote{This would occur with any
choice of the form $M(x, y) = \Phi(y) - \Phi(x)$ for a monotonic function
$\Phi()$.}. Specifically,
\begin{align}
M_1(x, y) + M_1(y, z)
    &= y^{1/3} - x^{1/3} + z^{1/3} - y^{1/3} \\
    &= z^{1/3} - x^{1/3} \\
    &= M_1(x, z).
\end{align}
This may not be entirely desirable. Genuinely popular claims on LBRY tend
to receive several smaller supports, rather than one large one, and intuitively
this ought to help. A trending algorithm should show what's trending,
as well as what's recently been boosted.

\subsection{Softened Change in LBC Amount}
Instead of applying the power before the subtraction, we could apply the
subtraction first --- i.e., we could use
\begin{align}
M_2(\ell_t, \ell_{t+1}) &= \left(\ell_{t+1} - \ell_t\right)^{1/3}.
\end{align}
The subscript 2 distinguishes this choice from $M_1()$ in the previous
subsection.
This would make several small supports more effective than one large support,
in terms of the trending score, a potentially desirable feature:
\begin{align}
M_2(x, y) + M_2(y, z)
    &= (y - x)^{1/3} + (z - y)^{1/3} \\
    &> (z - x)^{1/3}.
\end{align}
The inequality follows from (I don't know --- Jensen's inequality perhaps).
However, in numerical experiments, we found that this effect was too strong,
and whales could attain huge heights in trending score by splitting up, say,
100,000 LBC, into several supports.

\subsection{Interpolation}
The two ideas $M_1()$ and $M_2()$ can be merged into a compromise third option,
as follows:
\begin{align}
M_3(\ell_t, \ell_{t+1}) &= M_1(\ell_t, \ell_{t+1})^\alpha
                           M_2(\ell_t, \ell_{t+1})^{1 - \alpha}.
\end{align}
This is linear interpolation of $\log M$ values, or alternatively, it is
`geometric interpolation'. If $\alpha$ is close to 1 then the rule is more
like $M_1$, if it is close to zero it is more like $M_2$.

What we actually settled on was to have $\alpha \approx 0.5$ when we're dealing
with low LBC values, and $\alpha = 0.85$ when dealing with higher LBC values.
There is a narrow range over which $\alpha$ is between these two values:
\begin{align}
\alpha &= \left\{
        \begin{array}{lr}
            0.5, & \ell_2 \leq 50,\\
            0.01\ell_2, & 50 < \ell_2 \leq 85,\\
            0.85, & \textnormal{otherwise}.
        \end{array}
        \right.
\end{align}
This completes the specification of the spike mass function.
